Functional diversity improves tropical forest resilience to disturbance

### 2018-02-20 draft.Rmd
Draft 0.1.9002 logging removal, different disturbances & discussion

---
title: Draft
author: "Sylvain Schmitt sylvain.schmitt@agroparistech.fr"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2:
    includes:
      before_body: doc_prefix.tex
      in_header: preamble.tex
    number_sections: false
    toc: false
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float: yes
csl: /home/sylvain/Documents/Bibliography/csl/mee.csl
bibliography: /home/sylvain/Documents/Bibliography/library.bib
link-citations: yes
---

@@

**To be written (but eventually re-use [master-thesis one](https://sylvainschmitt.github.io/master-thesis/resume-et-abstract.html))**

@@

Tropical forests disturbances, through deforestation and degradation, account for 8.26 billion of tons of carbon dioxide emissions per year [@Pearson2017]. It represents the third source of greenhouse gas. Besides deforestation has been rightly the focus of worldwide attention, degradation has been less studied and quantified through tropics. Degradation from forest has been estimated to represents 10 time the deforestation area [@Herold2011] and represents 20% of greenhouse gas emissions in the Brazilian Amazon [@Asner2005]. Degradation has met numerous definitions [@Simula2009], but can be defined as the result from a disturbance event that induced a modification of the forest ecosystem reducing ecosystem services while keeping the ecosystem as a forest, contrary to deforestation.

Sustainable forest management in the tropics (i.e. managed selective harvesting of timber) has been widely promoted internationally to combat tropical deforestation and other forms of degradation [@Zimmerman2012]. Currently logging from tropical forests accounts for one eighth of global timber production [@Blaser2011] and is still increasing. Most tropical timber production originates from selective logging, the targeted harvesting of timber from commercial species in a single cutting cycle [@Martin2015].

On the other hand, tropical rainforests have fascinated ecologists due to their outstanding diversity [@connell_diversity_1978]. Effectively tropical forests host over half of the Earth's biodiversity [@Scheffers2012]. High biodiversity from tropical rainforests is the source of many ecosystem functions. Among others, tropical forests play a key role in biogeochemical cycles, including carbon storage [@Lewis2004]. Ecosystem functions from tropical forests support numerous ecosystem services, such as timber production and climate regulation.

But several authors argue that selective logging represents a major threat to biodiversity [@Carreno-Rocabado2012; @DeAvila2015; @Gibson2013; @Martin2015; @Zimmerman2012], challenging the sustainable definition from current selective logging. @DeAvila2015 found that additional sylvicultural practices and logging damages had stronger negative effect on diversity than logging itself. @Carreno-Rocabado2012 found no direct effect of medium intensity logging on species diversity but a community drift towards 'fast' acquisitive species [@reich_world-wide_2014]. However, even the short term impact on biodiversity is unclear, as sometimes short term increases in diversity can occur at low sylvicultural intensities [@Martin2015]. We consequently need to assess both short and long term impacts of selective logging on tropical forest ecosystems to implement better sylvicultural practices in order to reach sustainability.

The question of selective logging impact on tropical forest can be directly related to the emerging field of biodiversity and ecosystem functioning [@Loreau2000]. Tropical forest outstanding biodiversity will be both a factor and a result of forest ecosystem response to logging disturbance. And forest ecosystem response to logging disturbance will directly modify ecosystem functioning in both short and long term. Consequently assessing selective logging effect on tropical forest linking diversity and ecosystem functioning seems an obvious and promising way [@Loreau2010].

Negative short term impacts of selective logging have been assessed [@Carreno-Rocabado2012; @DeAvila2015;  but see @Martin2015]. Much less is known about the long term impact [@Osazuwa-Peters2015]. The main reason is the difficulty to conduct long term empirical study [but see @Herault2010], which can be completed by the use of forest simulators [@Huth2004; @Khler2004; @Ruger2008; @Tietjen2006]. Individual-based models of forest dynamics present the perfect framework to develop such joint biodiversity-ecosystem approaches [@Marechaux2017]. Individual-based models describe forest accumulating carbon through time, assessing tree growth, or releasing carbon through gap opening [@Bugmann2001]. Up to several dozens of different Plant Functional Types (PFTs) are generally defined and models can sometimes be fully spatially explicit [@Pacala1996]. Recently, the forest growth simulator TROLL [@Chave1999], an individual-based and spatially explicit forest model, was developed to introduce recent advances in plant ecophysiology. TROLL model relates physiological processes to species-specific functional traits [@Marechaux2017]. Consequently, TROLL model allow to simulate fully a neotropical forest biodiversity to study biodiversity-ecosystem functioning link response to logging disturbance.

@@

In the present study, we focused on mechanisms involved in the relationship between biodiversity and forest ecosystem resilience. We used a simulation approach using TROLL model to assess long term processes for different types and levels of disturbances. We first assessed diversity effect on forest resilience of structure and functioning using numerous indices of taxonomic and functional diversities. Then, we measured biodiversity net effect, partitioned into complementarity and selection effects, on resilience for several ecosystem metrics.

@@

We decided to explore the resilience of forest ecosystem following a generic disturbance. Disturbances impacting forest ecosystems are diverse as ... Disturbance module inside TROLL was designed in the simplest way in order to relate the ecosystem answer to volume loss without any individuals nor species targeting. For a given iteration $disturb_{iter}$, individuals are picked randomly with a uniform law on the number of trees. Selected individuals are then removed without triggering a treefall to avoid any side effect. The operation is repeated until the disturbance result in a defined lost basal area ($disturb_{intensity}$ in % of BA).

@@

```{r gIeq, echo=FALSE, fig.cap='Ecosystem resilience after 600 years with taxonomic and functional diversity for different levels of disturbance. Cummulative integral from ecosystem distance to forest structure equilibrium after 600 years was represented against functional diversity [FDiv, @villeger_new_2008] for different level of disturbance (25, 50 and 75% of total basal area); dot shapes represents the species richness whereas dot color represents functional evenness [FEve, @villeger_new_2008].', cache=TRUE, fig.width=12}
Ieq %>% 
  left_join(., doe, by = 'sim') %>% 
  filter(time == 600) %>% 
  ggplot(aes(x = FDiv, y = Ieq.structure, label = sim)) + 
  geom_smooth(method = 'lm', alpha = 0.3, colour= 'grey') +
  geom_point(aes(shape = richness, color = FEve)) +
  facet_wrap(~disturbance, scales = "free", labeller = 'label_both') + 
  xlab('Functional diversity (FDiv)') +
  ylab('Cummulative integral from ecosystem 
       distance to forest structure equilibrium after 600 years') +
  scale_shape_discrete('Species\nrichness\n(SR)') +
  scale_color_continuous('Functional\nevenness\n(FEve)')
```

@@

```{r gBE, echo=FALSE, fig.cap='Resilience of complementarity and selection effects. Complementarity effect (CE) and selection effect (SE) where normalized by control net effect (NEc), thus measuring their resilience over time.', cache=TRUE, fig.width=12}
BEresilience %>% 
  left_join(., doe, by = 'sim') %>%
  filter(richness == 125) %>% 
  filter(BE %in% c('SE', 'CE')) %>% 
  group_by(time, BE, disturbance, richness) %>%
  summarise(qt1 =quantile(agb, probs=0.05),
            qt2 =quantile(agb, probs=0.75),
            mean=mean(agb),
            n=n()) %>% 
  ggplot(aes(x = time)) +
  geom_hline(yintercept = 1, color = 'red') +
  geom_hline(yintercept = 0, color = 'blue') +
  geom_ribbon(aes(ymin = qt1, ymax = qt2, fill = BE), alpha = 0.3) +
  geom_line(aes(y = mean, color = BE)) +
  facet_grid(richness ~ disturbance, labeller = 'label_both') +
  xlab('Time (years)') +
  ylab('Recovery of aboveground biomass net effect')
```

@@

In the limit of the model, we were able to show that diversity improved tropical forest resilience. More particularly, functional diversity and evenness are key components of diversity in forest recovery after disturbance. Moreover, we found that complementarity between species was insuring forest recovery in forest succession start with facilitation before more productive species dominate the forest and insure recovery. Our results advocates for a sustainable harvesting of tropical forests through an increased resilience due to high diversity. But this conclusion should met a sustainable definition of selective logging following @Zimmerman2012. Because if the harvesting is not sustainable negative feedback will slowly diminish diversity and its benefits for forest resilience, resulting in forest degradation.

@@

Our results validated the hypothesis of a significative positive relationship between forest resilience and functional diversity and evenness ($p < 0.01$ and $p < 0.05$ respectively). Thus we were able to show that diversity improve forest resilience. More particularly, functional diversity [volume of the functional space occupied by the community, @villeger_new_2008] seems a major aspect of resilience if its strengthened by a high functional evenness [regularity of the distribution of abundance in this volume, @villeger_new_2008]. Indeed, high functional diversity needs evenness in order to better answer disturbance, if not the diversity is masked by ecosystem dominant species. Those results confirms the review of @Diaz2001 advocating for underevaluated importance of plant functional diversity in ecosystem processes. Additionally, the role of evenness  confirm the review of @Zhang2012 who highlighted the role of evenness in productivity and thus resilience following our hypothesis. Finally, species and functional richness are not directly increasing resilience. But increased species and functional richness will increase chance for high functional diversity through the sampling effect [@Loreau1998]. An higher sampling of regional species pool will allow a greater chance to pick more functionally diverse species.

@@

We found that complementary effect was insuring forest resilience in the beginning of forest successions. We interpreted this results as the consequence of facilitation processes. As the only resource currently simulated by TROLL is the light, the main facilitation will be light shading of post-pioneer species by pioneer species in disturbance gaps. Our results confirm the study of @Morin2011 who also highlighted the importance of facilitation through light shading in forest resilience. But the complementarity effect is reducing through time, to let the selection effect insure forest resilience due to an enforced dominance of more productive species. The diminishing of complementarity effect is attributed to competitive selection though time in gaps succession. But the study scale matters as shown by @Chisholm2013. Here we look at processes to a 16 ha scale. We do not have topography reducing micro-environment effects, but @Chisholm2013 results suggest that complementarity will be stronger at smaller scale and could explain its low value after forest recovery. Finally, our findings confirm results of @Tobner2016 realized on experimental forests of 4 years in low diverse forest of Canada. They also found that selection effect was greater than complementarity in most of the cases. In the case of selective logging the complementarity effect will thus be the major effect between two cutting cycles due to the cycle length (several decades in Guyana shield). High forest diversity with important species complementarity through increased functional diversity is thus an advantage for forest recovery between two cutting cycles in order to maintain both productive ecosystem and sustainable management.

@@

We used closed forest system simulated with TROLL forest model to evaluate the role and mechanisms of biodiversity in tropical forest resilience to disturbance. We found that diversity improved forest resilience, similarly to productivity [@Liang2016]. Additionally we found that complementarity between species was insuring forest recovery in forest succession start with facilitation, before more productive species dominate the forest and insure forest recovery. Our results advocate for sustainable selective logging against monoculture stand: high diversity will increase forest resilience and thus improve logging cycles. Even if monoculture stand are naturalized they will not reach mature forest diversity, and consequently they will not reach its natural high resilience. But on the other hand selective logging in tropical forest needs to be sustainable, if not the diversity will slowly decrease after each cycles degrading forest ecosystem functions, and resilience will decrease due to negative feedback. @Zimmerman2012 criticized the state of sylviculture in tropics, suggesting that "we have not been able to reconcile these opposing biological and economic forces". Still, they advocates for a possibility of sustainable tropical logging, already existing in small-scale, that will need proper funding from the international community. This view correspond to the high resilience of tropical forest hyperdiverse systems, we were able to show, and let hope for a future sustainable selective logging in tropical forests.

>>>>

---
title: Functional diversity improves tropical forest resilience to disturbance
author: "Sylvain Schmitt, Bruno Hérault, et al"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2:
    includes:
      before_body: ./template/doc_prefix.tex
      in_header: ./template/preamble.tex
    number_sections: false
    toc: false
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float: yes
linestretch: 1.5
csl: /home/sylvain/Documents/Bibliography/csl/mee.csl
bibliography: /home/sylvain/Documents/Bibliography/library.bib
link-citations: yes
---

@@

**To be written**

@@

Natural disturbance trough treefall is a main factor of forest dynamic [@Veblen1992]. Still, climate and human activities are also at the origin of forest disturbances, with huge and underevaluated impacts. Human-induced forest disturbances will result either in forest degradation or in deforestation. Degradation has met numerous definitions [@Simula2009], but can be defined as the result from a disturbance event that induced a modification of the forest ecosystem reducing ecosystem services while keeping the ecosystem as a forest, contrary to deforestation. Degradation can be the result of forest logging and management, whereas deforestation can be realized through total afforestation or fragmentation resulting in land use change. @Pearson2017 encompassed selective timber harvest in native forests, woodfuel harvest and fire in tropical forest degradation. Tropical forests disturbances, through deforestation and degradation, account for 8.26 billion of tons of carbon dioxide emissions per year [@Pearson2017]. It represents the third source of greenhouse gas. Besides deforestation has been rightly the focus of worldwide attention, degradation has been less studied and quantified through tropics. Degradation from forest has been estimated to represents 10 time the deforestation area [@Herold2011] and represents 20% of greenhouse gas emissions in the Brazilian Amazon [@Asner2005]. In addition to the direct impact of human on forests, a more insidious disturbance is climate change [@Herault2016].

Regarding deforestation, the ecosystem response to land use change is almost absent due to the negative feedback of the new ecosystem state after the tipping point was reached. Consequently, ecosystem response to disturbance should be studied in the different contexts of forest degradation. Concerning timber and woodfuel harvest, @DeAvila2015 found that additional sylvicultural practices and logging damages had stronger negative effect on diversity than logging itself. @Carreno-Rocabado2012 found no direct effect of medium intensity logging on species diversity but a community drift towards 'fast' acquisitive species [@reich_world-wide_2014]. However, even the short term impact on biodiversity is unclear, as sometimes short term increases in diversity can occur at low sylvicultural intensities [@Martin2015]. Forest fires have been shown to increase pioneers abundance, and if recurrent, to impoverish the forests in the eastern Amazon [@Cochrane1999a]. Finally, climate change effect on tropical areas results in a significant risk of precipitation decrease [@Davidson2012]. In addition, intense drought events has been shown to result in huge mortality rates and biomass loss [@Phillips2009].

Tropical rainforests have fascinated ecologists due to their outstanding diversity [@connell_diversity_1978]. Effectively tropical forests host over half of the Earth's biodiversity [@Scheffers2012]. High biodiversity from tropical rainforests is the source of many ecosystem functions. Among others, tropical forests play a key role in biogeochemical cycles, including carbon storage [@Lewis2004]. Ecosystem functions from tropical forests support numerous ecosystem services, such as timber production and climate regulation.

The question of disturbance impact on tropical forest can be directly related to the emerging field of biodiversity and ecosystem functioning [@Loreau2000]. Tropical forest outstanding biodiversity will be both a factor and a result of forest ecosystem response to disturbance. And forest ecosystem response to disturbance will directly modify ecosystem functioning in both short and long term. Consequently assessing disturbance effect on tropical forest linking diversity and ecosystem functioning seems an obvious and promising way [@Loreau2010].

Negative short term impacts of disturbances have been assessed for logging, fires and droughts ; but much less is known about the long term impact [@Osazuwa-Peters2015]. The main reason is the difficulty to conduct long term empirical study [but see @Herault2010], which can be completed by the use of forest simulators [@Huth2004; @Khler2004; @Ruger2008; @Tietjen2006]. Individual-based models of forest dynamics present the perfect framework to develop such joint biodiversity-ecosystem approaches [@Marechaux2017]. Individual-based models describe forest accumulating carbon through time, assessing tree growth, or releasing carbon through gap opening [@Bugmann2001]. Up to several dozens of different Plant Functional Types (PFTs) are generally defined and models can sometimes be fully spatially explicit [@Pacala1996]. Recently, the forest growth simulator TROLL [@Chave1999], an individual-based and spatially explicit forest model, was developed to introduce recent advances in plant ecophysiology. TROLL model relates physiological processes to species-specific functional traits [@Marechaux2017]. Consequently, TROLL model allow to simulate fully a neotropical forest biodiversity to study biodiversity-ecosystem functioning link response to a disturbance.

@@

In the present study, we focused on mechanisms involved in the relationship between biodiversity and forest ecosystem resilience. We used a simulation approach using TROLL model to assess long term processes for different levels of disturbances. We first assessed diversity effect on forest resilience of structure and functioning using numerous indices of taxonomic and functional diversities. Then, we measured biodiversity net effect, partitioned into complementarity and selection effects, on resilience for several ecosystem metrics.

@@

We decided to explore the resilience of forest ecosystem following a generic disturbance. Disturbance module inside TROLL was designed in the simplest way in order to relate the ecosystem answer to volume loss without any individuals nor species targeting. For a given iteration $disturb_{iter}$, individuals are picked randomly with a uniform law on the number of trees. Selected individuals are then removed without triggering a treefall to avoid any side effect. The operation is repeated until the disturbance result in a defined lost basal area ($disturb_{intensity}$ in % of BA).

@@

```{r gIeq, echo=FALSE, fig.cap='Ecosystem resilience after 600 years with taxonomic and functional diversity for different levels of disturbance. Cummulative integral from ecosystem distance to forest structure equilibrium after 600 years was represented against functional diversity [FDiv, @villeger_new_2008] for different level of disturbance (25, 50 and 75% of total basal area); dot shapes represents the species richness whereas dot color represents functional evenness [FEve, @villeger_new_2008]. Solid lines and grey areas show fitted relationships with their mean and variance of the form $Ieq_600 = FDiv + \\epsilon$.', cache=TRUE, fig.width=12}
Ieq %>% 
  left_join(., doe, by = 'sim') %>% 
  filter(time == 600) %>% 
  ggplot(aes(x = FDiv, y = Ieq.structure, label = sim)) + 
  geom_smooth(method = 'lm', alpha = 0.3, colour= 'grey') +
  geom_point(aes(shape = richness, color = FEve)) +
  facet_wrap(~disturbance, scales = "free", labeller = 'label_both') + 
  xlab('Functional diversity (FDiv)') +
  ylab('Cummulative integral from ecosystem 
       distance to forest structure equilibrium after 600 years') +
  scale_shape_discrete('Species\nrichness\n(SR)') +
  scale_color_continuous('Functional\nevenness\n(FEve)')
```

@@

```{r gBE, echo=FALSE, fig.cap='Resilience of complementarity and selection effects. Complementarity effect (CE) and selection effect (SE) where normalized by control net effect (NEc), thus measuring their resilience over time. Solid lines represents the mean value whereas areas represents the interval of values among simulations.', cache=TRUE, fig.width=12}
BEresilience %>% 
  left_join(., doe, by = 'sim') %>%
  filter(richness == 125) %>% 
  filter(BE %in% c('SE', 'CE')) %>% 
  group_by(time, BE, disturbance, richness) %>%
  summarise(qt1 =quantile(agb, probs=0.05),
            qt2 =quantile(agb, probs=0.75),
            mean=mean(agb),
            n=n()) %>% 
  ggplot(aes(x = time)) +
  geom_hline(yintercept = 1, color = 'red') +
  geom_hline(yintercept = 0, color = 'blue') +
  geom_ribbon(aes(ymin = qt1, ymax = qt2, fill = BE), alpha = 0.3) +
  geom_line(aes(y = mean, color = BE)) +
  facet_grid(richness ~ disturbance, labeller = 'label_both') +
  xlab('Time (years)') +
  ylab('Recovery of aboveground biomass net effect')
```

@@

In the limit of the model, we were able to show that diversity improved tropical forest resilience. More particularly, functional diversity and evenness are key components of diversity in forest recovery after disturbance. Moreover, we found that complementarity between species was insuring forest recovery in forest succession start with facilitation before more productive species dominate the forest and insure recovery. We thus advocate the view that both diversity theory (with niche complementarity and facilitation) and mass ratio theory (represented by selection) represents the mechanisms at the origin of functional diversity effect on ecosystem functions depending on the spatial scale and the time since the disturbance event.

@@

Our results validated the hypothesis of a significative positive relationship between forest resilience and functional diversity and evenness ($p < 0.01$ and $p < 0.05$ respectively). Thus we were able to show that diversity improve forest resilience. More particularly, functional diversity [volume of the functional space occupied by the community, @villeger_new_2008] seems a major aspect of resilience if its strengthened by a high functional evenness [regularity of the distribution of abundance in this volume, @villeger_new_2008]. Indeed, high functional diversity needs evenness in order to better answer disturbance, if not the diversity is masked by ecosystem dominant species. Those results confirms the review of @Diaz2001 advocating for underevaluated importance of plant functional diversity in ecosystem processes. Additionally, the role of evenness  confirm the review of @Zhang2012 who highlighted the role of evenness in productivity and thus resilience following our hypothesis. Finally, species and functional richness are not directly increasing resilience, thus supporting the idea that species diversity is often an inadequate surrogate of functional diversity [@Diaz2001; but see @Poorter2015]. Still, increased species and functional richness will increase chance for high functional diversity through the sampling effect [@Loreau1998]. An higher sampling of regional species pool will allow a greater chance to pick more functionally diverse species.

@@

We found that complementary effect was insuring forest resilience in the beginning of forest successions. We interpreted this results as the consequence of facilitation processes. As the only resource currently simulated by TROLL is the light, the main facilitation will be light shading of post-pioneer species by pioneer species in disturbance gaps. Our results confirm the study of @Morin2011 who also highlighted the importance of facilitation through light shading in forest resilience. But the complementarity effect is reducing through time, to let the selection effect insure forest resilience due to an enforced dominance of more productive species. The diminishing of complementarity effect is attributed to competitive selection though time in gaps succession.

In addition, the study scale matters as shown by @Chisholm2013 and @Sullivan2017. Here we look at processes to a 16-ha scale. On the other hand, our simulation does not include topography resulting in the absence of microenvironmental variation affecting both water and nutrient availability. And studies suggest that complementarity will be stronger at smaller scale [@Chisholm2013; @Sullivan2017], within plot itself (0.4-ha scale). This could explain the low value of complementarity after forest recovery. On the opposite, @Sullivan2017 suggested that environment at larger scale could obscure the positive relationship between functional diversity and ecosystem functions and explain why the relation is detected at 0.4-ha scale and not at 1-ha scale. Consequently, the absence of topography and environmental heterogeneity in our simulations could be the reason why we still observed an effect to a 16-ha scale, strengthening @Sullivan2017 interpretation.

Our results follows several studies advocating for multiple mechanisms through which functional diversity affect ecosystem functions. Complementarity effect encompass niche complementarity theory (or diversity theory), which has been shown to indirectly improve forest biomass and coarse wood productivity in subtropical forest [@Chiang2016]. Similarly, complementarity has been shown of primary importance in grassland functioning [@Cardinale2007; @Isbell2011]. On the other hand, selection effect can be related to the mass ratio hypothesis, which proposes that ecosystem functions are regulated by dominant species of the community [@Grime1998]. And mass ratio theory has been shown to primarily influence ecosystem functions with biomass accumulation (aboveground and shoots), productivity (coarse wood or aboveground), water use, and light interception for both grassland and subtropical forests [@Chiang2016; @Mokany2008]. Consequently, concerning mechanisms responsible for the positive effect of functional diversity on tropical forest functions after disturbance, we also advocate for the interplay of both niche complementarity and mass ratio theories, depending on time since the disturbance event and the spatial scale.

Finally, our findings confirm results from experimental studies. @Williams2017 also found experimentally an overyielding due to complementarity effect based on light and spatial complementarity with tree crown of young trees. In addition, @Tobner2016 studied 4-years stand in low diverse forests of Canada. They also found that selection effect was greater than complementarity in most of the cases, promoting the idea that mass ratio theory is the primary mechanisms involved in functional diversity effect on ecosystem functions [@Mokany2008].

@@

We used closed forest system simulated with TROLL forest model to evaluate the role and mechanisms of biodiversity in tropical forest resilience to disturbance. We found that diversity improved forest resilience, similarly to productivity [@Liang2016; @Poorter2015]. Additionally we found that complementarity between species was insuring forest recovery in forest succession start with facilitation, before more productive species dominate the forest and insure forest recovery.
